Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-09-05 paper/acknowledgements.tex
somechanges

This work was partly developed at the KITP conference 'Better stars, better
planets'.

@@

This work made use of the gaia-kepler.fun crossmatch database created by Megan
Bedell

>>>>

This work was partly developed at the 2019 KITP conference 'Better stars,
better planets'.
Parts of this project are based on ideas explored at the Gaia sprints at the
Flatiron Institute in New York City, 2016 and MPIA, Heidelberg, 2017.

This work made use of the gaia-kepler.fun crossmatch database created by Megan
Bedell.

@@

### 2019-09-05 paper/conclusion.tex
moving comments to comments
somechanges

We found that old groups of K dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
This indicates that either the \citep{angus2019} period-color relation is
wrong, or that mass-dependent dynamical heating is affecting this data set.
We presented three pieces of evidence that suggest mass-dependent heating
cannot be responsible for all of the increasing velocity dispersion.

\begin{itemize}

\item{
        No strong evidence for mass-dependent heating has been found among M, L
and T-type stars.
These low-mass stars live longer than the current age of the Universe and
        therefore do not suffer from a mass-age degeneracy
        \citep{faherty2009}.
}

\item{
        Literature measurements of the (\vz) AVR exponent are slightly higher
than the value measured from our data set, but calculated using more massive
stars.
Mass-dependent dynamical heating should however produce a {\it smaller} AVR
        exponent for higher-mass stars.
We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
        subject to different selection biases than those used to calculate
        AVRs in the literature \citep{holmberg2009, aumer2009, yu2018}, so
        drawing direct comparisons between them does not provide a reliable
        test.
}

\item{
        The differences in the (\vb) AVR exponent between the hottest and
coolest K dwarfs in our sample are extremely large, spanning values between
0.4 and 0.75, yet the difference in mass is less than a factor of two.
Such a large spread in AVR exponents cannot be explained by mass-dependent
        heating alone.
}

\end{itemize}

\racomment{Words about the significance of an inverted period-color relation.}

>>>>

We examined the gyrochronology relations using the velocity dispersions of
field stars with measured rotation periods.
We found that old groups of cool dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
This implies that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-color/\teff\ relation for old stars.
It appears that the period-color/\teff\ relation changes shape over time, \ie\
the period-color/\teff\ relation and the period-age relation are
non-separable.
At ages younger than $\sim$1 Gyr, rotation period is anti-correlated with
\teff: cooler stars spin more slowly than hotter stars of the same age.
However, at old ages the opposite is true: cooler stars spin more rapidly than
hotter stars of the same age.
The period-color/\teff\ relation appears to start flattening out after $\sim$1
Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
older than the period gap which is located at a gyrochronal age of around
$\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
This gap may represent a significant transitional epoch in the magnetic
behavior of stars.
Finally, when velocity dispersion is interpreted as an age proxy, it appears
that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
suggests that lower-mass stars remain active for longer and their rotation
periods can be measured at older ages.
We outlined a number of scenarios which could provide alternative
explainations for these observations, including incorrect dust corrections for
the lowest-mass stars and an excess of companions increasing the velocity
dispersion for these stars.

If the period-color/\teff\ relation does invert at old ages as our results
suggest, this would be a paradigm shift for gyrochronology.
Stellar spin-down rate is thought to be directly tied to magnetic field
strength, and the deeper convection zones of cooler stars generate stronger
magnetic fields which {\it should} lead to more efficient angular momentum
loss.
However, the micro- and macro-physics of stellar structure and evolution and
magnetic dynamo models are extremely complicated and a lot is still unknown
about the magnetic behavior of stars.
Observations like these can provide useful constraints for physical models,
and may help to reveal new physical processes at work in stars like our own
Sun and other planet hosts.

### 2019-09-05 paper/discussion.tex
somechanges

\section{Discussion}

What if the M dwarfs just have larger velocity dispersion because they're less
massive?

>>>>

\section{Discussion}
\label{sec:discussion}

\subsection{The period gap}
\label{sec:period_gap}

The origin of the rotation period gap, first identified
by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
\ref{fig:dispersion_period_teff} still remains a mystery.
This gap can be seen as an under-density of points between the 0.7-1.0 and
1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
of constant gyrochronal age of around 1.1 Gyr \citep[according to the
gyrochronology relation of][]{angus2019}, as shown in figure
\ref{fig:dispersion_period_teff}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018}, a rapid change in magnetic field structure
\citep{reinhold2019}, and erroneous rotation period measurements that are
incorrect by a factor of two \citep{koen2018}.
The latter explanation can be ruled out because stars below the gap have
smaller velocity dispersions than the stars above the gap, indicating that
they are kinematically younger \citep{mcquillan2013, davenport2018}, as
evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
velocity dispersion increases with \teff.
The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
dispersion as the hottest stars in the age range above which indicates that
the period-\teff\ relations are {\it flat} at these rotation periods.
This is also visible in figure \ref{fig:dispersion_period_teff}.
Below the gap, velocity dispersion within a given period range appears to {\it
decrease} with decreasing temperature.
The opposite appears to be true above the gap.
It could be that the gap is positioned at a significant Rossby number/age at
which stellar magnetic dynamos go through a transition.
Perhaps before the age of 1.1 Gyr, or at Rossby numbers less than ???,
magnetic braking is more efficient for stars with deeper convection zones.
Once stars reach this critical age or Rossby number their magnetic fields
undergo some radical transition, which produces the gap in the rotation
period-\teff\ plane.
After this transition, magnetic braking efficiency no longer increases with
decreasing mass.
Of course, it may be a coincidence that the gyrochronology relations seem to
only flatten off above the period gap and we lack a sufficient quantity of
data to do more than speculate here.
New rotation periods from the \ktwo\ and \tess\ missions may be able to
validate or rule out this hypothesis in the future.

In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
are faster than expected; their rotational evolution appears to have stalled,
and they have similar rotation periods to the 650 Myr Praesepe cluster
\citep{curtis2019}.
The rotation periods of the K dwarfs in this cluster are plotted in figure
\ref{fig:dispersion_period_teff}.
Although NGC 6811's G dwarfs fall on the 1.1 Gyr gyrochronology model, the K
dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
K dwarfs lie below it.
This cluster may be the `missing link' that connects two epoch of stellar
spin-down.
However, figure \ref{fig:age_cut} indicates that the Praesepe gyrochronology
model is, on average, a good model for stars younger than $\sim$1 Gyr. This
model has a very different shape to the NGC 6811 cluster.
This would suggest that stars in the field do follow the period-color/\teff\
relation of Praesepe, at least up to around 1 Gyr.
We do not find evidence of the K dwarf stalling, which is observed in NGC
6811, in the field.
% , however without more observations of middle-aged open clusters it : an
% early stage where the period-\teff\ relation for cool dwarfs has a negative
% slope and a late stage where it has a positive slope.
% In this case the period gap may delineate the transition between these two
% regimes and is the point at which stellar magnetic dynamos likely undergo a
% dramatic structural shift at an age of $\sim$ 1.1 Gyr.

% \subsection{Caveats}
% \label{sec:caveats}

% There are some alternative explanations for the trends seen in figures
% \ref{fig:age_cut} and \ref{fig:}
% We list these potential issues below.
% \begin{itemize}
% \item{
% The selection function for these data is extremely complicated, etc.
% }
% \item{
% Redenning for the coolest stars.
% }
% \item{
% Planets/companions for the coolest stars.
% }
% \item{
% Subgiant contamination/weakened magnetic braking.
% }
% \end{itemize}

### 2019-09-05 paper/intro.tex
somechanges

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation, calibrated using the period-color relation of Praesepe (in
\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it a good cluster for calibrating the period-color
relation of stars at 650 Myrs.
This relation accurately describes the rotation periods of F and G stars up to
around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
rotation periods), but over-predicts the rotation periods of K dwarfs in the
1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is difficult to characterize\footnote{The
oldest M dwarfs with reliable ages and rotation periods are members of
Praesepe ($\sim$ 650 Myrs), so even less is known about their rotational
evolution.}.

The gyrochronal ages of the stars in our sample were calculated with the
\citep{angus2019} gyrochronology model, using de-reddened \Gaia\ \gcolor\
color and rotation periods reported by \mct.
These rotational ages are shown on the CMD in figure \ref{fig:age_gradient}.
The stars with old rotational ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of stars with old rotational ages are bluer
than \gcolor\ $\sim$ 1.5 dex.
This provides a first hint that the \citet{angus2019} gyrochronology relation
under-predicts the ages of low-mass stars.
There is no reason to expect the oldest stars in this sample to be the bluer
ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
active for longer, so should therefore have measurable rotation periods at
older ages.
Since the \citet{angus2019} gyrochronology model, which is based on the
period-color relation of Praesepe, does not predict old ages for any M dwarfs,
it is probably under-predicting the ages of some of these low-mass stars.
It is therefore likely that old M dwarfs rotate more rapidly than this
Praesepe-based model predicts.
In what follows we use a population-based stellar age indicator, velocity
dispersion, to further investigate the gyrochronology relations at old ages in
the field.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
or $v_z$), just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time \citep[\eg][]{nordstrom2004, holmberg2007,
holmberg2009, aumer2009, casagrande2011}.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
The age-velocity dispersion relations (AVRs) are still actively being
calibrated, so it is difficult to directly compare gyrochronal ages with
kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.

Vertical action is thought to be a better age indicator than vertical velocity
\citep{ting2019}, although still only a weak age indicator for individual
stars \citep{beane2018}, however both vertical action ($J_z$) and vertical
velocity (\vz/W) can only be calculated with full 6-dimensional position and
velocity information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close (although imperfect -- see section \ref{sec:results})
approximation to \vz.

This paper is laid out as follows: in section \ref{sec:the_data} we describe
our sample selection process and the methods used to calculated stellar
velocities.
In section \ref{sec:results} we use stellar kinematics to investigate the
relationship between rotation period, age and color/\teff\ in the field.
In section \ref{sec:mass-dependent-heating} we show that neither
mass-dependent heating nor the selection function is likely to have a strong
affect on our sample, and in section \ref{sec:period_gap} we discuss the
implications of our results.

>>>>

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation, calibrated using the period-color relation of Praesepe (in
\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it a good cluster for calibrating the period-color
relation of stars at 650 Myrs.
However, although this relation accurately describes the rotation periods of F
and G stars up to around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster
with available rotation periods), it over-predicts the rotation periods of K
dwarfs in the 1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is largely unknown.
The oldest M dwarfs with reliable ages and rotation periods are members of
Praesepe ($\sim$ 650 Myrs), so even {\it less} is known about their rotational
evolution.

In this paper, gyrochronal ages of cool field dwarfs in the \kepler\ field
were calculated with the \citep{angus2019} gyrochronology model, using
de-reddened \Gaia\ \gcolor\ color and rotation periods reported by \mct\ (we
explain our data selection process in section \ref{sec:the_data}).
These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
figure \ref{fig:age_gradient}.
The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
This provides the first hint that the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
There is no reason to expect the oldest stars in this sample to be the bluer
ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
active for longer, so should therefore have measurable rotation periods at
older ages.
Since the \citet{angus2019} gyrochronology model, which is based on the
period-color relation of Praesepe, predicts the oldest stars in this sample to
be K dwarfs, it is probably either under-predicting M dwarf ages or
over-predicting K dwarf ages.
% In other words, old M dwarfs rotate more rapidly or K dwarfs rotate more
% slowly than the model predicts.
In what follows, we use a population-based stellar age indicator, velocity
dispersion, to investigate the gyrochronology relations at old ages in
the field.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the galaxy with a low out-of-plane, or vertical, velocity ($W$, or \vz),
just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
On average, the vertical velocities of stars increase over time
\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
casagrande2011}.
Although the cause of dynamical heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% The age-velocity dispersion relations (AVRs) are still actively being

% Vertical action is thought to be a better age indicator than vertical
% velocity \citep{ting2019}, although still only a weak age indicator for
% individual stars \citep{beane2018}, however both vertical action ($J_z$) and
% vertical
Vertical velocity, \vz, can only be calculated with full 6-dimensional
position and velocity information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close (although imperfect -- see section \ref{sec:results})
approximation to \vz.
Because we use \vb\ rather than \vz\, it is difficult to directly compare
gyrochronal ages with kinematic ones using an age-velocity dispersion relation
(AVR).
However, regardless of direction, velocity dispersion is expected to
monotonically increase over time, and can therefore be used effectively to
{\it rank} groups of stars by age.

This paper is laid out as follows: in section \ref{sec:method} we describe our
sample selection process and the methods used to calculated stellar
velocities.
We also establish that \vb\ velocity dispersion, \sigmavb, can be used as an
age proxy by demonstrating that neither mass-dependent heating nor the
selection function is likely to have a strong affect on our sample.
In section \ref{sec:results} we use stellar kinematics to investigate the
relationship between rotation period, age and color/\teff\ in the field.
We show that the period-color relation of Praesepe is not applicable to
old stars in section \ref{sec:age_cut}, and reveal the true shape of the
period-color relations in section \ref{sec:the_reveal}.
We discuss a possible connection with the rotation period gap in section
\ref{sec:period_gap}.
